---
title: "Multi Linear Regression-RQ1"
output: pdf_document
date: "2022-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load Libraries
library(ggplot2)
library(psych)
library(corrplot)
library(dplyr)
library(table1)
library(boot)
library(MASS)
library(equatiomatic)
library(jtools)
library(car)
library(kableExtra)
```

```{r}
#Load the data
df1 <- read.csv("https://raw.githubusercontent.com/EricR401S/Modeling-Trends-for-Spotify-Songs/part2/archive/rq1.csv")
```

## Linear Regression Model

Our initial model contained all the 7 predictors selected in the data section above was put through a backward AIC selection to come up with the most pertinent predictors of popularity.

```{r}
linear1 <- lm(popularity ~ acousticness+danceability+energy+instrumentalness+ tempo+ loudness+speechiness , data = df1)
backward <- stepAIC(linear1, trace=FALSE, direction = 'backward')
summ(backward, digits = 4)
```
**Should we think of mean centering these variables?**

A centering attempt.

```{r}
#plot(backward)
df1$acousticness_centered <- scale(df1$acousticness, scale = FALSE)
df1$danceability_centered <- scale(df1$danceability, scale = FALSE)
df1$energy_centered <- scale(df1$energy, scale = FALSE)
df1$instrumentalness_centered <- scale(df1$instrumentalness, scale = FALSE)
df1$tempo_centered <- scale(df1$tempo, scale = FALSE)
df1$loudness_centered <- scale(df1$loudness, scale = FALSE)
df1$speechiness_centered <- scale(df1$speechiness, scale = FALSE)

scaled <- lm(popularity ~ acousticness_centered+danceability_centered+energy_centered+instrumentalness_centered+tempo_centered+loudness_centered+speechiness_centered, data = df1)
summ(scaled, digits = 7)
```
```{r}
confinterval_scaled <- confint(scaled, level = 0.95)
kable(confinterval_scaled, caption =  "conf interval table for scaled model") %>% kable_styling(position="center",latex_options = c("hold_position"))
```

**Acousticness** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average acoustiness for a song with a 39 popularity score is `r mean(df1$acousticness)`.We are 95% confident that (x1, x2) contains the true value for acoustiness and for every 1 *unit* increase in acoustiness of a song the average popularity score increases by approximately 1.8 holding all other factors constant.

**Danceability** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average danceability for a song with a 39 popularity score is `r mean(df1$danceability)`.We are 95% confident that (x1, x2) contains the true value for danceability and for every 1 *unit* increase in danceability of a song the average popularity score increases by approximately 11.7 holding all other factors constant. 

**Energy** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average energy for a song with a 39 popularity score is `r mean(df1$energy)`.We are 95% confident that (x1, x2) contains the true value for energy and for every 1 *unit* increase in energy of a song the average popularity score *increases by approximately 1.8 holding all other factors constant* 

**Instrumentalness** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average instrumentalness for a song with a 39 popularity score is `r mean(df1$instrumentalness)`.We are 95% confident that (x1, x2) contains the true value for instrumentalness and for every 1 *unit* increase in nstrumentalness of a song the average popularity score *increases by approximately 1.8 holding all other factors constant.* 

**Tempo** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average tempo for a song with a 39 popularity score is `r mean(df1$tempo)`.We are 95% confident that (x1, x2) contains the true value for tempoand for every 1 *unit* increase in tempo of a song the average popularity score decreases by approximately 0.007 holding all other factors constant.

**Loudness** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average loudness for a song with a 39 popularity score is `r mean(df1$loudness)`.We are 95% confident that (x1, x2) contains the true value for loudness and for every 1 *unit* increase in loudness of a song the average popularity score increases by approximately 1 holding all other factors constant. 

**Speechiness** is a significant predictor of a song's popularity score at the $\alpha = 0.05$ significance level ($p< 0.001$).The average speechiness for a song with a 39 popularity score is `r mean(df1$speechiness`.We are 95% confident that (x1, x2) contains the true value for speechiness and for every 1 *unit* increase in speechiness of a song the average popularity score increases by approximately 6.4 holding all other factors constant.


```{r}
#plot(backward)
#df$Duration_centered <- scale(df$Duration, scale = FALSE)
equatiomatic::extract_eq(scaled, use_coefs = TRUE)
```
##Assumptions

1. Linearity : Residual vs predictor graphh

2. Normal distribution of residuals: Normal Q-Q CURVE : Parses

3. Independence of error term: Residual vs fitted curve

4. homoscedasticity:( Equal variance of error terms) The Scale-Location graphs shows whether residuals are spread equally along the ranges of predictors. It’s good we see a horizontal line with equally (randomly) spread points. 


>Linearity: The residuals vs predictor plot 

```{r}
#Assumptions of model 1
#Linearity residuals vs duration
library(olsrr)
res <- resid(scaled)
plot(df1$acousticness,res)
plot(df1$danceability,res)
plot(df1$energy,res)
plot(df1$instrumentalness,res)
plot(df1$tempo,res)
plot(df1$loudness,res)
plot(df1$speechiness,res)
abline(0,0)
#Independence and equal variance of error terms
ols_plot_resid_fit(scaled)

```

> Independence of error terms : The points on the residual vs fitted plot are randomly spread 

```{r}
#Independence and equal variance of error terms
ols_plot_resid_fit(backward)
```

> Equal Variance of error terms : The points on the residual vs fitted plot are equally spread around the ab line(0,0) therefore we can conclude that there is homoscedasticity.

```{r}
#Independence and equal variance of error terms
ols_plot_resid_fit(backward)
```

> Normality of residuals: The points on the quantile-quantile plot are clustered along the 45 degree line so there are no clear violations of this assumption.

```{r}
ols_plot_resid_hist(backward)
#Normality of residuals
ols_plot_resid_qq(backward)
```

## Checking Multicollinearity

The model appears to have no issues with multicollinearity as seen below all the VIPO values are below 5 therefore the potential multicollinearity between energy and acoustiness is not there. 
```{r}
multicol <- vif(linear1)
kable(multicol, caption =  "VIF table") %>% kable_styling(position="center",latex_options = c("hold_position"))
summ(backward)
```


## Interpretation of Coeffcients

```{r}
confinterval <- confint(backward, level = 0.95)
kable(confinterval, caption =  "conf interval table") %>% kable_styling(position="center",latex_options = c("hold_position"))
```


## Outliers and Influential point - Cooks distance
```{r}
ols_plot_cooksd_bar(backward)
```

```{r}
ols_plot_resid_lev(backward)
```

```{r}
#ols_plot_resid_lev(backward[])
```

