---
title: "Untitled"
author: "Eric Rios"
date: "2022-11-11"
output: pdf_document
---

## R Markdown

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Load Libraries
library(ggplot2)
library(psych)
library(corrplot)
library(dplyr)
library(table1)
library(boot)
library(caret)
library(arm)
library(pROC)
library(e1071)
library(stargazer)
library(car)
```

```{r subset, echo = FALSE, message=FALSE, results = FALSE}
#read original data set
df <- read.csv("https://github.com/EricR401S/Modeling-Trends-for-Spotify-Songs/raw/main/archive/tracks.csv")
```

```{r datacleaning, echo = FALSE, message=FALSE, results = FALSE }
#Two columns of interest : Name of artist (if one wanted to highlight certain summary statistics), duration conversion (get minutes), the rest is perfect on its own. 

# Confirming the data types of the columns
sapply(df, class)

# Removing the brackets from the names of the artists
# This will facilitate summary statistics

df$artists<-gsub("]","",as.character(df$artists))
df$artists<-gsub("^.","",as.character(df$artists))

# New minute variable for our own use to simplify interpretation

df$duration_minutes <- df$duration_ms/(1000*60)


# Confirming that there are no missing data, except the artist name column

colSums(is.na(df))

# change "explicit" into binary factor
df$explicit_fac <- factor(df$explicit,
                         levels=c(0,1),
                         labels=c('Non-Explicit','Explicit'))

# make Date readable to R
# Emma and Eric need to revise this together, some records with Dates passed as NAs in R after running this code.
# That's the theory. It was single year dates.
df$release_year <- substr(df$release_date, 1, 4)
df$release_year <- as.integer(df$release_year)


# According to variable definitions, speechiness levels above 0.66 are speech tracks such as podcasts and poetries.
# They will be removed.

df0 <- df[df$speechiness <= 0.66,]
nrow(df) - nrow(df0) #22,598 records of speech tracks

# Examining records with a value of 0 for tempo
# A total of 328 records with 0 tempo were found, and most were tracks of rain sounds and introductions. 
# Some were, unfortunately, real songs.

sum(df0$tempo==0)

# 148 of those records are from the 2010s decade, our area of interest. 
tempo_0_subset <- df0[df0$tempo == 0,]
tempo_0_subset_2010s <- tempo_0_subset[grep('201[0-9].*', tempo_0_subset$release_date),]
nrow(tempo_0_subset_2010s)

# Removing records with a value of 0 for tempo

df0 <- df0[df0$tempo != 0,]
```

```{r, echo = FALSE, message=FALSE, results = FALSE }
#subset data
subset <- df0[grepl('201[0-9].*', df0$release_year),]
```

```{r , echo=FALSE,results="asis", header=FALSE, message=FALSE, warning=FALSE}
RQ1_relation <- c("popularity", "acousticness", "danceability", "energy", "instrumentalness", "tempo", "loudness", "speechiness")
df1 = subset[RQ1_relation]
RQ2_relation <- c("explicit_fac", "danceability", "energy", "speechiness","tempo", "release_year")
df2 = subset[RQ2_relation]
```

```{r}
train <- df2[df2$release_year < 2018,]
test <- df2[df2$release_year >= 2018,]
keeps <- c("explicit_fac", "danceability", "energy", "speechiness", "tempo")
test <- test[keeps]
train <- train[keeps]
```

```{r warning=FALSE, echo=FALSE, message = FALSE, results='asis'}
r2 <- glm(explicit_fac ~ ., data = train , family = binomial(link = logit))
test_fitted_0 <- predict(r2, newdata = test, type = "response") # Change Here
test_fitted <- ifelse(test_fitted_0 >= 0.5, "Explicit", "Non-Explicit") #Change here
Conf_mat3 <- confusionMatrix(as.factor(test_fitted), test$explicit_fac)
ConfMat3 <- as.data.frame.matrix(Conf_mat3$table)
Accuracy3 <- Conf_mat3$overall["Accuracy"]
stargazer(ConfMat3, summary = FALSE, type = "text", title = "Confusion Matrix")
stargazer(Accuracy3, type = "text", summary = FALSE)
invisible(roc(test$explicit_fac,test_fitted_0,plot=T,print.thres=0.5,legacy.axes=T, print.auc =T,col="red3", main = "ROC Curve -- Logistic Regression Model : Explicitness")) # Inclusion here

```


```{r}
library(car)
a <- vif(r2) 
stargazer(a, type = "text", summary = FALSE)
```
Second Research Question Portion

# Abstract: 
A few sentences describing the purpose of the analysis, the data, and key results


TBW



#Introduction: Provide more background on the data and research questions. Be sure to cite the data
and background information appropriately (APA style is fine)



TBW (Copy Paste with Citations)


# Methods
Describe the process you used to conduct analysis. This includes EDA and any relevant
data cleaning information (e.g., did you exclude missing values? If so, how many? Did you collapse categories for any variables?) Then describe the models you fit, and any changes you made to improve model fit (e.g., did you exclude any influential points? Did you do have to address multicollinearity issues? Did you transform any variables?). Also describe model diagnostics. The organization of this section may depend on your particular dataset/analysis, but you may want to break it into subsections such as “Data,” “Models,” and “Model assessment.” Note that you do not present any results in this section.

P.S. I noticed the results have to be moved to the results section. I'm just writing them together and splitting them later. 

For this analysis, we chose to use logistic regression to model the relationship between explicitness and its predictors. This type of regression model is used for variables that have only two values from which to choose (cite). In this case, a song is either explicit or non-explicit, two values. The second step was to divide our data into training and tests sets, where the former and latter are comprised of songs from 2010-2017 and 2018-19 respectively. Then the process concludes with three final steps, assessing the significance of the predictors in the model as well as their insights, estimating multicollinearity, and evaluating the model's accuracy. For the first, the coefficients of our predictors indicate the odds that our song may be explicit (cite). For the second, we verify that our variables are not influencing each other too much because they are assumed to be independent of each other. For the third, the model generated from the train set is used to predict the likelihood of a song being explicit or non-explicit in the test set, which is the final step in logistic regression (cite). We then gauge the effectiveness of our model by observing the general accuracy values and the Area Under the Curve (AUC) value, another measure of how well the model is predicting the negatives and positives in a study (cite). 

The interpretations : 

As discussed previously, a low P-value is a sign that our predictor variable has an association with our outcome variable. Shown in our table below, our model's predictors are all statistically significant in predicitng a song's explicitness. 

In terms of interpreting the coefficients, holding all other variables constant, the odds of a song being explicit...

1. increase by 8.40 times for every unit increase in the danceability predictor.

2. increase by 2.18 times for every unit increase in the energy predictor. 

3. increase by 10,855 times for every unit increase in the speechiness predictor.

4. decrease by 0.94 times for every unit increase in the tempo predictor.

The results here have logical sense. What is most impressive is how danceability is strongly related to a track's explicitness. In other words, very danceable records are more likely have explicit lyrics. Energy was a bit weaker in this regard, but also implied a similar relationship. As for speechiness, its extremely high odds are best understood by thinking "the more words a song has, the more likely it is to have some explicit content". Lastly, tempo's odds indicate that, as a song becomes slower, it will be less likely to have explicit lyrics. 



```{r}
stargazer(exp(cbind(OR = coef(r2))), type = "latex", report = ("vcsp*"),header = FALSE, single.row = TRUE, digits = 4, no.space = TRUE, column.sep.width = "3pt", title = "Odds Ratio : Logistic Regression Model for Explicitness")
```

---------------------Might need a section here for the binned residuals------------------------------


------------------------------------------/----------------------------------------------------------


In good practice, the VIF values will also be verified. They are hardly above 1, which means the variables are independent.

VIF table





Lastly, we look at the final and most important piece, the accuracy of our model on the test set. Shown below are the ROC curve and the confusion matrix, as well as the overall accuracy values. One of our challenges was addressing the small number of explicit songs in our data, where it might hurt model accuracy. That concern was proven true when our model failed at correctly classifying 3,937 explicit songs. It had an overall, high accuracy of 81%, a sensitivity of 97.9% and a specificity of 14%. The sensitivity value signifies that our model is efficient at identifying non-explicit songs, while our specificity value demonstrates that our model is lacking when attempting to identify an explicit song. Also, the AUC value was a 79% success rate of distinguishing negatives and positives. 

Lorna's Explanation - Asking for verification

Since our model is weak at identifying explicit songs, what does this mean for our study? Firstly, there was a smaller distribution of explicit songs; however, this pattern is indicative of music in the real world because the majority of songs are non-explicit. Secondly, more or better predictors may have been needed for this particular research question. Thirdly, is it critical for this model to predict these songs' explicitness accurately? No. Spotify likely employs the use of speech recognition (natural language processing) models to flag the explicitness of songs, not statistical methods. However, since the intent of the research question is predictive, these results do prove that our model is insufficient for predicting the explicitness of songs, but very successful when classifying the non-explicitness of songs.    



ROC plot here

Confusion Matrix table with labels (True Negatives, True Positives, we just need to know which one is which)

Table (with accuracy and AUC values for test)



# Results
Here you should present results for all aspects of the analysis. The structure of this section should mirror the structure of the methods section. For example, you can start with a few key EDA results (e.g., a table of descriptive statistics), then present model results, then address assessment. This is the section where you will primarily refer to tables and figures. You should have at least 1 figure for each research question that illustrates a key result of the analysis.






# Conclusion
Describe the key takeaways from your analysis, limitations, and future work that can be done to advance knowledge in this area.



TBW